### 2009-12-14 docs/marcello-dissertation/thesis/chapter8-experimental-results-correct-behavior-and-performance.tex
Reviews of misspelling and references to pages, Section, etc.
- Chapter 7: Fixes on "Chapter" to chapter
- Chapter 8: Subversion Conflict line removal, and other fixes
- Appendix: Removing "on page x" for the image on the same page.

git-svn-id: svn://localhost/netbeams/branches/marcello/persistence@706 559edde2-a4d4-44e0-96fc-4035e40d7396

This chapter details the experiment setup designed to evaluate the
proposed implementation, showing the measurement results collected from the
simulation log and based on the user experience while reusing the collected
data. The pertinent discussion of this work is divided up into different
sections, each of them analyzing different aspects related to data persistence
for NetBEAMS, among others.

\section{Experiment Setup}

@@

In order to evaluate the DSP Data Persistence component, experiments were
conducted using the DSP Mouse Actions component, a software sensor simulator
that captures the actions performed by the mouse pointer.

@@

Different artifacts were created to perform the different experiments regarding
the CRUD operations. The ``Create'' operation' was developed in Java, isolating
the services provided by the class DSPMongoCRUDService (Listing
\ref{file:dsp-mongo-service}). It is responsible for the connection handling
with mongoDB, as well as the insertion operation from NetBEAMS. As described
\cite{netbeams-dsp-architecture}, the current version of NetBEAMS supports the
data collection for the YSI Sonde device \cite{YSI-Sonde}, among others used
for demo purposes. After evaluating the database with the YSI Sonde data
handler, the use of the demo component DSP Mouse Actions was used. It captures
the mouse pointer activities performed over a window, for example, mouse clicks
and mouse grag over coordinates x and y.

In order to exercise mongoDB with the same data collected from NetBEAMS, the
existing Java class TestSondeData (Listing
\ref{file:random-ysi-data-generator}), was refactored to provide a random
instance generator for the YSI Data handler implemented in the Java class
SondeDataType (Listing \ref{file:main-ysi-data-handler}). When the Insert
function is executed, the created instances are transferred from the JVM to
the mongoDB, and stored in the collection ``SondeDataContainer''. Specific
details about the execution of mongoDB are depicted in Section
\ref{sec:mongodb-deployment}. Finally, the implementation of the experiment
in the Java class DSPMessageToMongoDBExperiment (Listing
\ref{file:experiment-dsp-java-executor}) was responsible for managing the
creation and insertion of any workload composed by instances of the class
SondeDataType in a bulk manner.

Similarly, the support for the Retrieve, Update and Delete operations against
the inserted workload were implemented using Javascript, as shown in Listing
\ref{file:experiment-scenarios}, respectively by the function calls ``find'',
``update'' and ``remove'' over the collection ``SondeDataContainer''.
Similarly, in order to implement a customizable export system that converts
from mongoDB into the format OpenDAP, a simpler prototype script was developed
in Python, as shown in Listing \ref{file:experiment-export-python}. As a result
of the execution of this script, the exported artifact has the same format as
described in Listing \ref{file:rtc-ysi-opendap}.

@@

\begin{itemize}
  \item (R0) Insert a volume of collected data from sensors using NetBEAMS
  service compatible with 1 year of observations;
  \item (R1) Find all documents whose observations were collected between two
  different dates, say the days between December 8 and 15 of 2009;
  \item (R2) Find all documents whose observations were collected from a given
  known sensor device such as one whose IP address is equals to "192.168.0.102";
  \item (R3) Find all documents whose observations contained specific values
  read from the environment, say Salinity equals to 0.01 and the water temperature
  equals to 46.47;
  \item (U1) Update all the documents whose observations were collected during
  the day of December 2nd, 2009, by adding a tag = ``oil spill'';
  \item (D1) Remove all the documents whose observations were collected on
  December 7th, 2009.
  \item (R4) Export the produced data of the week to CSV files, using the same
  tabular format sequence used in the files distributed by SF-BEAMS using the 
  the OPeNDAP format, as in Listing \ref{file:rtc-ysi-opendap}.
\end{itemize}

@@

The first set of measurements collected were the ones regarding the use case
R1, shown in Listing \ref{file:dsp-mongo-service}, defined as the insertion of
instances of the collected data from the sensor devices using Java. The
document follows the specification of the keys designed in Section
\ref{sec:dsp-persistence-data-model}. The construction of the keys
and respective values were done using a Hash-like object ``BasicDBObject''
such as the ones shown in the method ``buildKeySegment''. Similarly, the
creation of database indexes was performed using the same approach, as shown in
the Java method ``setupMetadataIndexes''. Then, the insertion of the
documents was performed on the service method call
``insertPersistentUnitMessageContents'', which delimits the bulk creation using
the transaction management mechanism over the collection. In this way,
different insert times were collected from the logs of the different
experiments as summarized on Table 7.1.

@@

In general, the user experience of mongoDB is related the methods used to
access and modify the state of the database using a different approach when
compared to traditional Relational databases. Since most of the methods to
create, access and modify the database were performed using the programming
language abstraction of objects and method calls, the transition from the use
of a relational data model is very smooth. The creation of data was the first
contact with mongoDB in terms of data creation in Java. Later, different
methods of data access were used after the randomly generated data was
inserted into mongoDB: using the shell commands mongo and ``mongoexport'',
command-line execution developed in Python, through the REST Web Services API,
and using a web application, written in the Ruby Programming Language
\cite{ruby}, providing the view over the collected data using a web browser.
More details about the execution of these commands during the experiments are
shown in Section \ref{sec:mongodb-user-experience}.

The primary method used to access the mongoDB database is using the shell
command. Once the user requests access to the database, he or she can issue an
operation over the collection of documents using regular the abstraction
programming language, as the shown in Listing \ref{file:experiment-scenarios}.
The CRUD operations are available to the user and can be used at any time to
any database collection. Aggregation operations such as to count the returned
result set is also provided via programming languages abstraction.

In order to export the collected data to other data formats, the  command
``mongoexport'' and the developed export script written in Python were used.
The ``mongoexport'', exports the collected data from the database using the
native support to comma-delimited values. As shown in Listing
\ref{file:mongodb-export-command}, the actual execution describe how mongo is
querying the data from the database. As a result, the total number of documents
exported is shown in on line 8. In addition to the regular command that
exports the entire database, the user can provide a query string during the
execution of the command in order to filter the dataset returned and,
consequently, the number of documents exported. Similar to the use of the
default export capabilities of mongoDB, a script written in Python to export
data from mongoDB was written as shown in Listing
\ref{file:experiment-export-python}. It uses the package ``pymongo'' to create
a comma-delimited file with the same header and line format as shown in Listing
\ref{file:experiment-export-python-results}. The results can be compared with
the example the output of the HTTP Response body of Listing
\ref{file:rtc-ysi-opendap}, requested directly to the OPeNDAP server of RTC.
Therefore, the experiment is capable of reproducing theat the OPeNDAP can.

While the access to the database using the shell command gives full access to 
the CRUD operations, the access to the data was also performed using the REST
Web Services API, as shown in Listing \ref{cmd:mongo-rest-request}. As one can
see, the IP address and the default port number of the mongoDB REST middleware
are provided. Then, the REST operation uses the name of the database and
the collection to correlate with the ones in the database system. Thus,
querying the collections from mongoDB can be performed using the HTTP Request
Parameter variables, by adding the prefix ``filter\underline{ }'' before the
name of the key of the document. The given example shows the query performed
in the collected data for the key ``observation.Conductivity'' with the value
``104.5''. Similarly, the HTTP Request variable ``limit'' is used to reduce
the number of documents returned by the query. The HTTP Response from mongoDB
returns the result using the JSON format, similar to the ones printed in the
command-shell.

@@

The actual execution of the solution was performed using the demo Mouse
Actions, which consists of a software sensor that perceives any mouse action in
a window area. The experiment consisted of moving the mouse around a white
window, where each action is captured and temporarily saved until a threshold is
reached, making the component to send the messages to the persistence layer, so
that it can be saved on mongoDB. Figure \ref{fig:experiment-mouseactions-dsp}
shows a screenshot of the moment when the message is sent to mongoDB, being
verified through the HTTP REST Interface using a web browser
\ref{sec:dsp-mongodb-rest-ws}.

\begin{figure}[h]
  \centering
  \includegraphics[scale=0.4]{../diagrams/experiment-mouseactions-dsp}
  \caption{Experiment showing the mongoDB persisting messages from the DSP
  Mouse Actions Demo}
  \label{fig:experiment-mouseactions-dsp}
\end{figure}

@@

Any database system covers the purpose of the sensor data for NetBEAMS, which
is Data Archival, along with the type of storage for the data as an External
Storage. After selecting mongoDB, the development of the persistence layer met
the requirements of archiving data for different types of sensor devices from
the DSP Platform such as the YSI Data Handler and the Mouse Actions demo data.
As it is shown in Section \ref{sec:exp-measurements}, the collected data is
archived in files using the BSON format \cite{bson}, a binary representation
used to store instances of the document described in Listing
\ref{file:mongodb-ysi-data-format}. Along with the purpose of Data Archival,
the use of mongoDB also met the specification of External Storage, since the
data persisted in the file system, as well as Data-Centric for running mongoDB
with different  distributed mongo shards (cluster nodes).

SF-BEAMS uses Tabular Data model, a schema-less data model that represents the
collected data using comma-delimited files, without a database system to manage
those files, just the OPeNDAP middleware. In contrast, the data persistence
method proposed by this work suggests mongoDB as a schema-less database system
to manage the data collected using NetBEAMS. As described in Chapter 5, the
schema-less data model can provide better support to Data Provenance because
the data model allows the change of records without affecting the
others. In this way, the metadata used by the implementation clearly describes
the three fundamental questions regarding the collected data from sensor
devices, as described in Section \ref{sec:dsp-persistence-data-model}. The 
identity of the data is given by the key ``message\underline{ }id'', in a way
that it uniquely identifies the data produced by the sensor identified by the
key ``sensor.ip\underline{ }address''. Thus, the message can be tracked
back to the sensor that produced the data for different purposes such as
tracking the data producer. Then, time dimensions were used in order to
track two different aspects of the data collection: the valid time can be used
to correlate observations to the exact moment it occurred, while the
transaction time can be used to verify when the data was collected, and make
decisions such as deleting the instances of data collected in the previous two
days, etc. Similarly, the use of the key ``sensor.location'' can give the exact
location of the data, since researchers can be interested in finding data
based on a specific location. The most important aspect of the schema-less data
model is that the observed data is added to the key ``observation'', as any
sensor device carries its set of attributes of key-value pairs. A good example
of the applicability of the schema-less model is the scalability regarding
changes to the data structure. For example, if an existing sensor device is
flushed with a new version of the firmware\footnote{Primary machine
instructions that run in the hardware}, changing the format of data types or
adding new values, the following instances of collected data may very
from the previous one by simply adding the new additions of data. For this
reason, the DSP Data Persistence gives administrators the ability to choose
which fields to persist by declaring it using a bootstrap message for the
component, as shown in the parameter ``YSI\underline{ }DESIRED\underline{
}PROPERTIES'' of Listing \ref{file:dsp-data-persistence-bootstrap.xml}.

In order to access the collected data, the implementation used different APIs
provided by the mongoDB community. The Centralized Query Mechanism was
satisfied by the use of those APIs, which also fulfills the requirements
defined by NetBEAMS. For example, the accesses of the database using Python or
Java were based on the needs and the use of the languages for different
purposes. In this way, the implementation of the persistence model captures
the instances of any known data collected from NetBEAMS and saves it using the
document pattern as shown in Listing \ref{file:mongodb-ysi-data-format}. Any
programming language used was capable of accessing the same collected data by
their own.

In the end, the implementation of the data persistence for NetBEAMS was
developed in both Single and Distributed Systems. The use of regular single
server execution provided enough computational power for a compatible volume of
data when compared to the one produced from the SF-BEAMS sensor network. The
implementation suggested by this work fulfill the needs of a persistence system
for a Small volume of data being created at the same time in a Single Server.
Furthermore, mongoDB also provided support for a Distributed System, when the
use of database shards, in forms of different mongoDB nodes, was used as a 
clustered environment.

@@

The performance to insert data into the database system using the Java driver
is documented by mongoDB as the fastest approach, providing persistence
thousands of documents per second. This capacity is more than necessary for
NetBEAMS, which can use the Single System environment. Improvements over the
response time can be achieved through the implementation of a Data-Centric
approach on a Distributed Database environment, or clusters. In this approach,
the inserted data is stored in different servers based on a partition key.
Given that less data is concentrated in each partition, the centralized query
mechanism used is faster, and consequently improves the performance as
described in the Data Location taxonomy in Chapter 3. In this way, the
infrastructural performance during the inserts suggests that persisting data
collected from the NetBEAMS sensors using a clustered environment is an option
when the single server capacity is reached.

@@

The development of the scenarios were clearly written in different ways, such
as the ones defined in Listing \ref{file:experiment-scenarios}, which
implements the use cases for data Retrieval, Update and Deletion operations. The
main advantage of this approach is that users do not use the SQL language to extract
data from the database, but use method calls defined for each the
abovementioned operations. For example, based on the description of the
collected data by keys and values shown in Listing
\ref{file:mongodb-ysi-data-format}, different use cases can evaluate documents
in this structure. Usually researchers may search the collected data by a
given range of dates, for their regular activities of data analysis. The
scenario implemented by R1 in the above cited listing shows it clearly.
Similarly, the search for data in a given sensor device is expressed by the
search for the IP address for them on the use cased defined by R2. Finally,
the search for the raw data obtained by the sensor devices is used from the
key ``observation'', as shown by the R3 use case implementation.

@@

Commonly during scientific research, data is shared. Researchers can choose
the use of spreadsheets to analyze collected data with simpler tools than
database systems or programming language. For this reason, OPeNDAP is used in
the community to share data using a comma-delimited format. In this way, the
implementation of the use cases for R4 was developed in two different ways.
First, the use of the export capabilities to a common-delimited values was
performed \ref{file:mongodb-export-command}, taking proximally 3.3 minutes to
export one year's worth of data for 5 YSI devices without indexing. Second, the
implementation of a custom export utility using Python, as shown in Listing
\ref{file:experiment-export-python} enabled the creation of the comma-delimited
files with the same format as the ones used by OPeNDAP. The comparison can be
done by analyzing the output of the OPeNDAP request from SF-BEAMS on Listing
\ref{file:rtc-ysi-opendap}, and the output provided from the script written in
Python, show in Listing \ref{file:experiment-export-python-results}.
Furthermore, considering research teams with skills in Web Services, an easy
ubiquitous integration with mongoDB can be achieved by using the REST Web
Services APIs provided by mongoDB, still in alpha version. A practical application of
this API was the created Apache Futon4mongoDB web application, used to give
user friendlier access to the collected data as shown in 
\ref{fig:view-collections-instance-browser-futondb}.

@@

\begin{figure}[!h]
  \centering
  \includegraphics[scale=0.65]{../diagrams/taxonomy-data-model-modified}
  \caption{Data Model Taxonomy}
  \label{fig:taxonomy-data-model-modified}
\end{figure}

@@

This work can be used to evaluate data persisten in sensor networks in
different ways. The use of the data persistence taxonomies through this work
and guided both the evaluation and implementation of the data model using
mongoDB. In order to use the guidelines proposed by this work, one may first
determine the properties of the sensor network in question. Considering that
sensor devices produce streams of non-described data, the use of key-value
pairs data model did not require changes throughout different application
layers.

Focusing on the different layers of an application, consider the different
multi-tier application approaches where each layer implements a functionality
of the system such as the Model-View-Controler (MVC) \cite{sw-architectures}.
Usually, the use of traditional database systems and object-programming 
languages require ``data plumbing'' from the table relations from
the database system to the object instances. In order to implement all the
CRUD operations to a relational database without an ORM \cite{orm} offering,
the implementation of the data model potentially uses the Structured-Query
Language (SQL) for the problem of data plumbing, what results in a
tightly-coupled implementation. On the other hand, mongoDB eliminates the use
of an additional query language because it uses declarative approach on top of
the programming language drivers it support, decreasing the complexity of the
application.

@@

Particularly speaking about the learning curve, different situations
contributed to the success of the first experiments. The first problem to be
solved was the selection of the data model to be used during the technology
selection. Since relational database systems are the ``traditional'' way to
provide data persitence, the investigations started by creating a relational
model described in Chapter 5. Then, the technique using KVP data models in
relational databases were attempted without success. Only after reviewing the
technical article \cite{db-is-rdbs-dommed}, suggested the advisor of this 
work, the search for a database system that uses the KVP data model was
considered, and finally successfully implemented.

@@

Physical memory is limited and, for this reason, it may not be enough to run
the experiments for total workloads. In this case, the Java Virtual Memory was
totally consumed by one of the revisions of the experiments setup. Therefore, 
it was learned that large amount of data loads require different techniques to
manage data in memory accordingly, such as creating different object pools as
temporary caches for objects whose internal state are similar to others in the
JVM. Also, operations that involve writing in disk, such as during the bulk
``Create'' operations, may lead to different database performance. First, the
experiment seemed to stop inserting objects in one shard while it ran for over
2 minutes without inserting any documents. The reason was that the partition
was ``/home'', completely full, leading to the problem. The mongos cluster
header did not realize that it was the time to switch to another shard.
Similarly, any global ``Retreave'' operation on a cluster blocks all the
shards. For this reason, dividing the write from the read load was necessary in
order to decouple these two operations.

Also related to the difficulties of this work, mongoDB is not yet stable, and
therefore, harder to implement. The main production version of mongoDB is
is on version 1.1.4 at this time, while the shards support is still in alpha
version. For instance, different work-arounds were performed to have shards
experimented as described, but the execution was not successful.

To sum up, this chapter described the experimental results of the
implementation of a data persistence component for NetBEAMS, showing the
experiment results, the measurement results and finally the discussion about
the solution in general. The following section brings the conclusions and
future works for this report.

>>>>

This chapter details the experiment setup designed to evaluate the proposed
implementation, showing the measurement results collected from the simulation
log and based on the user experience while reusing the collected data. The
pertinent discussion of this work is divided up into different sections, each
of them analyzing different aspects related to data persistence for NetBEAMS.
Section \ref{sec:exp-setup} introduces the experiments setup and methods used
for the evaluation such as the artifacts used, the definition of the workload
to be used during the evaluation of the planned scenarios of evaluation. Then,
Section \ref{sec:exp-measurements} shows the measurement results of the
experiment detailing the operations performed by each of the scenarios and
user experience, to finally proceed with the final analysis and discussions.

\section{Experiment Setup}
\label{sec:exp-setup}

@@

@@

Different artifacts were created to perform the different experiments regarding
the CRUD operations. The ``Create'' operation' was developed in Java, isolating
the services provided by the class DSPMongoCRUDService (Listing
\ref{file:dsp-mongo-service}, page \pageref{file:dsp-mongo-service}). It is
responsible for the connection handling with mongoDB, as well as the insertion
operation from NetBEAMS. As described \cite{netbeams-dsp-architecture}, the
current version of NetBEAMS supports the data collection for the YSI Sonde
device \cite{YSI-Sonde}, among others used for demo purposes.

In order to exercise mongoDB with the same data collected from NetBEAMS, the
existing Java class TestSondeData (Listing
\ref{file:random-ysi-data-generator}, page
\pageref{file:random-ysi-data-generator}), was refactored to provide a random
instance generator for the YSI Data handler implemented in the Java class
SondeDataType (Listing \ref{file:main-ysi-data-handler} on page
\pageref{file:main-ysi-data-handler}). When the Insert function is executed,
the created instances are transferred from the JVM to the mongoDB, and stored
in the collection ``SondeDataContainer''. Specific details about the execution
of mongoDB are depicted in Section \ref{sec:mongodb-deployment}. Finally, the
implementation of the experiment in the Java class
DSPMessageToMongoDBExperiment (Listing
\ref{file:experiment-dsp-java-executor}, page
\pageref{file:experiment-dsp-java-executor}) was responsible for managing the
creation and insertion of any workload composed by instances of the class
SondeDataType in a bulk manner.

Similarly, the support for the Retrieve, Update and Delete operations against
the inserted workload were implemented using Javascript, as shown in Listing
\ref{file:experiment-scenarios} on page \pageref{file:experiment-scenarios},
respectively by the function calls ``find'', ``update'' and ``remove'' over
the collection ``SondeDataContainer''. Similarly, in order to implement a
customizable export system that converts from mongoDB into the format OpenDAP,
a simpler prototype script was developed in Python, as shown in Listing
\ref{file:experiment-export-python} on page
\pageref{file:experiment-export-python}. As a result of the execution of this
script, the exported artifact has the same format as described in Listing
\ref{file:rtc-ysi-opendap}, page \pageref{file:rtc-ysi-opendap}.

@@

\begin{itemize}
  \item (R0) Insert a volume of collected data from sensors using NetBEAMS
  service compatible with 1 year of observations;
  \item (R1) Find all documents whose observations were collected between two
  different dates, say the days between December 8 and 15 of 2009;
  \item (R2) Find all documents whose observations were collected from a given
  known sensor device such as one whose IP address is equals to "192.168.0.102";
  \item (R3) Find all documents whose observations contained specific values
  read from the environment, say Salinity equals to 0.01 and the water temperature
  equals to 46.47;
  \item (U1) Update all the documents whose observations were collected during
  the day of December 2nd, 2009, by adding a tag = ``oil spill'';
  \item (D1) Remove all the documents whose observations were collected on
  December 7th, 2009.
  \item (R4) Export the produced data of the week to CSV files, using the same
  tabular format sequence used in the files distributed by SF-BEAMS using the 
  the OPeNDAP format, as in Listing \ref{file:rtc-ysi-opendap} on page
  \pageref{file:rtc-ysi-opendap}.
\end{itemize}

@@

The first set of measurements collected were the ones regarding the use case
R1, shown in Listing \ref{file:dsp-mongo-service} page
\pageref{file:dsp-mongo-service}, defined as the insertion of instances of the
collected data from the sensor devices using Java. The document follows the
specification of the keys designed in Section
\ref{sec:dsp-persistence-data-model}. The construction of the keys and
respective values were done using a Hash-like object ``BasicDBObject'' such as
the ones shown in the method ``buildKeySegment''. Similarly, the creation of
database indexes was performed using the same approach, as shown in the Java
method ``setupMetadataIndexes''. Then, the insertion of the documents was
performed on the service method call ``insertPersistentUnitMessageContents'',
which delimits the bulk creation using the transaction management mechanism
over the collection. In this way, different insert times were collected from
the logs of the different experiments as summarized on Table 7.1.

@@

In general, the user experience of mongoDB is related to the methods used to
access and modify the state of the database using a different approach when
compared to traditional Relational databases. Since most of the methods to
create, access and modify the database were performed using the programming
language abstraction of objects and method calls, the transition from the use
of a relational data model is very easy. The creation of data was the first
contact with mongoDB in terms of data creation in Java. Later, different
methods of data access were used after the randomly generated data was
inserted into mongoDB: using the shell commands mongo and ``mongoexport'',
command-line execution developed in Python, through the REST Web Services API,
and using a web application, written in the Ruby Programming Language
\cite{ruby}, providing the view over the collected data using a web browser.
More details about the execution of these commands during the experiments are
shown in Section \ref{sec:mongodb-user-experience}.

The primary method used to access the mongoDB database is using the shell
command. Once the user requests access to the database, he or she can issue an
operation over the collection of documents using regular the abstraction
programming language, as the shown in Listing \ref{file:experiment-scenarios}
on page \pageref{file:experiment-scenarios}. The CRUD operations are available
to the user and can be used at any time to any database collection.
Aggregation operations such as to count the returned result set is also
provided via programming languages abstraction.

In order to export the collected data to other data formats, the  command
``mongoexport'' and the developed export script written in Python were used.
The ``mongoexport'', exports the collected data from the database using the
native support to comma-delimited values. As shown in Listing
\ref{file:mongodb-export-command}, page \pageref{file:mongodb-export-command},
the actual execution describe how mongo is querying the data from the
database. As a result, the total number of documents exported is shown in on
line 8. In addition to the regular command that exports the entire database,
the user can provide a query string during the execution of the command in
order to filter the dataset returned and, consequently, the number of
documents exported. Similar to the use of the default export capabilities of
mongoDB, a script written in Python to export data from mongoDB was written as
shown in Listing \ref{file:experiment-export-python}, page
\pageref{file:experiment-export-python}. It uses the package ``pymongo'' to
create a comma-delimited file with the same header and line format as shown in
Listing \ref{file:experiment-export-python-results}, page
\pageref{file:experiment-export-python-results}. The results can be compared
with the example the output of the HTTP Response body of Listing
\ref{file:rtc-ysi-opendap} on page \pageref{file:rtc-ysi-opendap}, requested
directly to the OPeNDAP server of RTC. Therefore, the experiment is capable of
reproducing the targeted OPeNDAP format used by the RTC server.

While the access to the database using the shell command gives full access to 
the CRUD operations, the access to the data was also performed using the REST
Web Services API, as shown in Listing \ref{cmd:mongo-rest-request} on page
\pageref{cmd:mongo-rest-request}. As one can see, the IP address and the
default port number of the mongoDB REST middleware are provided. Then, the
REST operation uses the name of the database and the collection to correlate
with the ones in the database system. Thus, querying the collections from
mongoDB can be performed using the HTTP Request Parameter variables, by adding
the prefix ``filter\underline{ }'' before the name of the key of the document.
The given example shows the query performed in the collected data for the key
``observation.Conductivity'' with the value ``104.5''. Similarly, the HTTP
Request variable ``limit'' is used to reduce the number of documents returned
by the query. The HTTP Response from mongoDB returns the result using the JSON
format, similar to the ones printed in the command-shell.

@@

@@

After selecting mongoDB as the best candidate for the persistence
infrastructure for NetBEAMS, the development of the persistence layer met the
requirements of archiving data for different types of sensor devices from the
DSP Platform such as the YSI Data Handler. As it is shown in Section
\ref{sec:exp-measurements}, the collected data is archived in files using the
BSON format \cite{bson}, a binary representation used to store instances of
the document described in Listing \ref{file:mongodb-ysi-data-format} on page
\pageref{file:mongodb-ysi-data-format}. Along with the purpose of Data
Archival, the use of mongoDB also met the specification of External Storage,
since the data persisted in the file system, as well as Data-Centric for
running mongoDB with different distributed mongo shards (cluster nodes).

SF-BEAMS uses Tabular Data model, a schema-less data model that represents the
collected data using comma-delimited files, without a database system to manage
those files, just the OPeNDAP middleware. In contrast, the data persistence
method proposed by this work suggests mongoDB as a schema-less database system
to manage the data collected using NetBEAMS. As described in Chapter 5, the
schema-less data model can provide better support to Data Provenance because
the data model allows the change of records without affecting the
others. In this way, the metadata used by the implementation clearly describes
the three fundamental questions regarding the collected data from sensor
devices, as described in Section \ref{sec:dsp-persistence-data-model}. The 
identity of the data is given by the key ``message\underline{ }id'', in a way
that it uniquely identifies the data produced. Thus, the message can be tracked
back to the sensor that produced the data for different purposes such as
tracking the data producer. Then, time dimensions were used in order to
track two different aspects of the data collection: the valid time can be used
to correlate observations to the exact moment it occurred, while the
transaction time can be used to verify when the data was collected, and make
decisions such as deleting the instances of data collected in the previous two
days, etc. Similarly, the use of the key ``sensor.location'' can give the exact
location of the data, since researchers can be interested in finding data
based on a specific location. The most important aspect of the schema-less data
model is that the observed data is added to the key ``observation'', as any
sensor device carries its set of attributes of key-value pairs. A good example
of the applicability of the schema-less model is the scalability regarding
changes to the data structure. For example, if an existing sensor device is
flushed with a new version of the firmware\footnote{Primary machine
instructions that run in the hardware}, changing the format of data types or
adding new values, the following instances of collected data may very
from the previous one by simply adding the new additions of data. For this
reason, the DSP Data Persistence gives administrators the ability to choose
which fields to persist by declaring it using a bootstrap message for the
component, as shown in the parameter ``YSI\underline{ }DESIRED\underline{
}PROPERTIES'' of Listing \ref{file:dsp-data-persistence-bootstrap.xml}, page
\pageref{file:dsp-data-persistence-bootstrap.xml}.

In order to access the collected data, the implementation used different APIs
provided by the mongoDB community. The Centralized Query Mechanism was
satisfied by the use of those APIs, which also fulfills the requirements
defined by NetBEAMS. For example, the accesses of the database using Python or
Java were based on the needs and the use of the languages for different
purposes. In this way, the implementation of the persistence model captures
the instances of any known data collected from NetBEAMS and saves it using the
document pattern as shown in Listing \ref{file:mongodb-ysi-data-format} on page
\pageref{file:mongodb-ysi-data-format}. Any programming language used was
capable of accessing the same collected data by their own.

In the end, the implementation of the data persistence for NetBEAMS was
developed in both Centralized and Distributed Systems. The use of regular single
server execution provided enough computational power for a compatible volume of
data when compared to the one produced from the SF-BEAMS sensor network. The
implementation suggested by this work fulfills the needs of a persistence system
for a Small volume of data being created at the same time in a Single Server.
Furthermore, mongoDB also provided support for a Distributed System, when the
use of database shards, in forms of different mongoDB nodes, was used as a 
clustered environment.

@@

The performance to insert data into the database system using the Java driver
is documented by mongoDB as the fastest approach, providing persistence for
thousands of documents per second. This capacity is more than adequate for
NetBEAMS, which can use the Single System environment. Improvements over the
response time can be achieved through the implementation of a Data-Centric
approach on a Distributed Database environment, or clusters. In this approach,
the inserted data is stored in different servers based on a partition key.
Given that less data is concentrated in each partition, the centralized query
mechanism used is faster, and consequently improves the performance as
described in the Data Location taxonomy in Chapter 3. In this way, the
infrastructural performance during the inserts suggests that persisting data
collected from the NetBEAMS sensors using a clustered environment is an option
when the single server capacity is reached.

@@

The development of the scenarios were clearly written in different ways, such
as the ones defined in Listing \ref{file:experiment-scenarios} on page
\pageref{file:experiment-scenarios}, which implements the use cases for data
Retrieval, Update and Deletion operations. The main advantage of this approach
is that users do not use the SQL language to extract data from the database,
but use method calls defined for each the abovementioned operations. For
example, based on the description of the collected data by keys and values
shown in Listing \ref{file:mongodb-ysi-data-format}, page
\pageref{file:mongodb-ysi-data-format}, different use cases can evaluate
documents in this structure. Usually researchers may search the collected data
by a given range of dates, for their regular activities of data analysis. The
scenario implemented by R1 in the above cited listing shows it clearly.
Similarly, the search for data in a given sensor device is expressed by the
search for the IP address for them on the use cased defined by R2. Finally,
the search for the raw data obtained by the sensor devices is used from the
key ``observation'', as shown by the R3 use case implementation.

@@

Commonly during scientific research, data is shared. Researchers can choose
the use of spreadsheets to analyze collected data with simpler tools than
database systems or programming language. For this reason, OPeNDAP is used in
the community to share data using a comma-delimited format. In this way, the
implementation of the use cases for R4 was developed in two different ways.
First, the use of the export capabilities to a common-delimited values was
performed (Listing \ref{file:mongodb-export-command}, page
\pageref{file:mongodb-export-command}), taking proximally 3.3 minutes to export
one year's worth of data for 5 YSI devices without indexing. Second, the
implementation of a custom export utility using Python, as shown in Listing
\ref{file:experiment-export-python}, page
\pageref{file:experiment-export-python}, enabled the creation of the
comma-delimited files with the same format as the ones used by OPeNDAP. The
comparison can be done by analyzing the output of the OPeNDAP request from
SF-BEAMS on Listing \ref{file:rtc-ysi-opendap}, page
\pageref{file:rtc-ysi-opendap}, and the output provided from the script
written in Python, show in Listing \ref{file:experiment-export-python-results}
page \pageref{file:experiment-export-python-results}. Furthermore, considering
research teams with skills in Web Services, an easy ubiquitous integration
with mongoDB can be achieved by using the REST Web Services APIs provided by
mongoDB, still in alpha version. A practical application of this API was the
created Apache Futon4mongoDB web application, used to give user friendlier
access to the collected data as shown in Figure
\ref{fig:view-collections-instance-browser-futondb}, on page
\pageref{fig:view-collections-instance-browser-futondb}.

@@

\begin{figure}[!h]
  \centering
  \includegraphics[scale=0.65]{../diagrams/taxonomy-data-model-modified}
  \caption{Updated Data Model Taxonomy, adding the KVP Data Model}
  \label{fig:taxonomy-data-model-modified}
\end{figure}

@@

This work can be used to evaluate data persistence in sensor networks in
different ways. The use of the data persistence taxonomies through this work
guided both the evaluation and implementation of the data model using mongoDB.
In order to use the guidelines proposed by this work, one may first determine
the properties of the sensor network in question. Considering that sensor
devices produce streams of non-described data, the use of key-value pairs data
model did not require changes throughout different application layers.

Focusing on the different layers of an application, consider the different
multi-tier application approaches where each layer implements a functionality
of the system such as the Model-View-Controler (MVC) \cite{sw-architectures}.
Usually, the use of traditional database systems and object-programming 
languages require ``data plumbing'' from the table relations from
the database system to the object instances. In order to implement all the
CRUD operations to a relational database without an ORM \cite{orm} offering,
the implementation of the data model potentially uses the Structured-Query
Language (SQL) for the problem of data plumbing, which results in a
tightly-coupled implementation. On the other hand, mongoDB eliminates the use
of an additional query language because it uses declarative approach on top of
the programming language drivers it support, decreasing the complexity of the
application.

@@

Particularly speaking about the learning curve, different situations
contributed to the success of the first experiments. The first problem to be
solved was the selection of the data model to be used during the technology
selection. Since relational database systems are the ``traditional'' way to
provide data persistence, the investigations started by creating a relational
model described in Chapter 5. Then, the technique using KVP data models in
relational databases were attempted without success. Only after reviewing the
technical article \cite{db-is-rdbs-dommed}, suggested the advisor of this 
work, the search for a database system that uses the KVP data model was
considered, and finally successfully implemented.

@@

Physical memory is limited and, for this reason, it may not be enough to run
the experiments for total workloads. In this case, the Java Virtual Memory was
totally consumed by one of the revisions of the experiments setup. Therefore, 
it was learned that large amount of data loads require different techniques to
manage data in memory accordingly, such as creating different object pools as
temporary caches for objects whose internal state are similar to others in the
JVM. Also, operations that involve writing in disk, such as during the bulk
``Create'' operations, may lead to different database performance. First, the
experiment seemed to stop inserting objects in one shard while it ran for over
2 minutes without inserting any documents. The reason was that the partition
was ``/home'', completely full, leading to the problem. The mongos cluster
header did not realize that it was the time to switch to another shard.
Similarly, any global ``Retreave'' operation on a cluster blocks all the
shards. For this reason, dividing the write from the read load was necessary
in order to decouple these two operation.

Also related to the difficulties of this work, mongoDB is not yet stable, and
therefore, harder to implement. The main production version of mongoDB
is on version 1.2 at this time, while the shards support is still in alpha
version. For instance, different work-arounds were performed to have shards
experimented as described, but the execution was not successful.

To sum up, this chapter described the experimental results of the
implementation of a data persistence component for NetBEAMS, showing the
experiment results, the measurement results and finally the discussion about
the solution in general. The following chapter brings the conclusions and
future works for this report.
